---
title: 'Time Series Analysis of Out-of-Service Buses'
author: 'UW DSSG Paratransit Group'
date: 'updated: Monday 31st August 2015'
output: html_document
---
# Time Series Analysis #

In the following analyses, we analyze the daily percentage of out-of-service buses for King County Metro (KCM). Buses become out-of-service generally after breaking down or if they are unable to arrive at their destinations. Predicting out-of-service buses is important from an operational perspective because KCM and their contractors can better manage their vehicle fleet by preparing adequate standby vehicles each day.

Our model uses the daily percentage of out-of-service buses relative to all buses operating that day (instead of raw counts of out-of-service buses). Using time series techniques, we analyzed almost 18 months (January 2014 to May 2015) of KCM transit data and concluded:
<ul>
<li>On average, 7.47% of buses are out-of-service each day</li>
<li>No deterministic trend over time</li>
<li>Some weekly seasonal variation</li>
<li>A snowstorm on February 9th 2014 coincided with the highest daily percentage of out-of service buses</li>
<li>Out-of-service bus percentages are likely stationary</li>
<li>Cross validation comparisons between ARMA models indicate an AR(1)MA(2) may be the best model to forecast future daily percentages of out-of-service buses</li>
</ul>
```{r, warning=FALSE, message=FALSE, include =FALSE}
rm(list=ls())
library(MASS)             
library(zoo)
library(timeDate)
library(tseries)
library(forecast)
library(RColorBrewer)      
library(TTR)
library(xts)
library(quantmod)  	   

# loading data
setwd('../data')
code8Buses<-read.csv("code8_counts.csv")
attach(code8Buses)
par(mfrow=c(1,1))
```

```{r}
# creating out-of-service buses variable (percentage)
pctCode8<-(oos_runs/total_runs)*100
code8Buses$pctCode8<-pctCode8
meanCode8<-summary(pctCode8)[4]
meanCode8

# determining peak out-of-service day
peakpctCode8<-code8Buses[which(code8Buses$pctCode8 ==max(code8Buses$pctCode8)),]
peakpctCode8

# peak day for out-of-service buses (35%) is due to snowstorm
# http://www.seattleweatherblog.com/snow/winter-wonderland-seattle-sees-biggest-february-snowfall-in-13-years/

# Plot - Daily Percentage of Out-of-Service Buses Plot
#png(filename = "Plot - Out-of-Service Buses.png")
plot(pctCode8 ,type="l",ylab="Percentage Out-of-Service Buses",xlab="Days", 
     main = "Daily Percentage of Out-of-Service Buses \n January 2014 - May 2015 ", axes = TRUE)
text(sprintf("Average Daily Percentage \n of Out-of-Service Buses \n %.02f percent", meanCode8),x = 140, y = 20, col="red",cex=0.7) 
abline(a=meanCode8,b=0, col = 'red',lty="dashed") # comparison line
abline(v=40,col = 'red',lty="dashed") # comparison line
text("Sunday, February 9th 2014", x = 140, y = 2, col="red",cex=0.7 )
#dev.off()

```

```{r, include = FALSE}
par(mfrow=c(1,1))
length(pctCode8)
timevar<-seq(1,length(pctCode8 ),1)
```

```{r}
# creating weekly time series of out-of-service bus percentages
ts.pctCode8 <- ts(pctCode8 ,1,frequency=7)
summary(ts.pctCode8)

# Plot - Out-of-Service Buses Time Series
#png(filename = "Plot - Out-of-Service Time Series.png")
plot(ts.pctCode8,type="l",col="black",ylab="Percentage of Out-of-Service Buses",xlab="Week", axes =FALSE, main = "Weekly Time Series of Out-of-Service Buses \n January 2014 - May 2015")
text(sprintf("Average Daily Percentage \n of Out-of-Service Buses \n %.02f percent", meanCode8),x = 140, y = 20, col="red",cex=0.7) 
axis(1,1:80)
axis(2,seq(0,36,2))
text(sprintf("Average Daily Percentage \n of Out-of-Service Buses \n %.02f percent", meanCode8),x = 20, y = 20, col="red",cex=0.7) 
abline(a=meanCode8,b=0, col = 'red',lty="dashed") # comparison line
abline(v=6.56,col = 'red',lty="dashed") # comparison line
text("Sunday, February 9th 2014", x = 20, y = 2, col="red",cex=0.7 )
#dev.off()
par(mfrow=c(1,1))

summary(lm(ts.pctCode8~timevar)) # timevar does not have deterministic trend
reg_residuals = residuals(lm(ts.pctCode8~timevar))
acf(reg_residuals, lag=7)

```

```{r, include = FALSE}
#if deterministic trend prsent, remove with the following loop

#ts.pctCode8.detrend<-rep(0,length(pctCode8 )) 
#for (i in 1:length(pctCode8 )){ 
#  ts.pctCode8.detrend[i] <- ts(ts.pctCode8[i] - lm(ts.pctCode8~timevar)$coefficients[2]*timevar[i]) #} 
#plot(ts.pctCode8.detrend,type="l", col="red",ylab="Out-of-Service Buses",xlab="Days")
#dev.off()
```

```{r}
# Plot - Time Series Decomposition
#png(filename = "Plot - Time Series Decomposition.png")
#ts.pctCode8.detrend<-ts(ts.pctCode8.detrend, frequency = 7)
ts.pctCode8.decomposition <- decompose(ts.pctCode8,type="additive")
plot(ts.pctCode8.decomposition)
#dev.off()
par(mfrow=c(1,1))

# data has some seasonalilty
# removing seasonal trend from percentage of out service buses 
ts.pctCode8.deseasonalized<-ts.pctCode8-ts.pctCode8.decomposition$seasonal 
meanCode8Deaseasonalized<-summary(ts.pctCode8.deseasonalized)[4]
meanCode8
meanCode8Deaseasonalized

# Plot - Out-of-Service Time Series (Deseasonal)
#png(filename = "Plot - Out-of-Service Time Series (Deseasonal).png")
plot(ts.pctCode8.deseasonalized,type="l",col="black",ylab="Out-of-Service Buses",xlab="Week", axes =FALSE, main = "Out-of-Service Buses \n without Seasonal Trend") 
axis(1,1:80)
axis(2,seq(0,36,2))
text(sprintf("Average Daily Percentage \n of Out-of-Service Buses \n %.02f percent", meanCode8),x = 20, y = 20, col="red",cex=0.7) 
abline(a=meanCode8Deaseasonalized,b=0, col = 'red',lty="dashed") # comparison line
abline(v=6.56,col = 'red',lty="dashed") # comparison line
text("Sunday, February 9th 2014", x = 20, y = 2, col="red",cex=0.7 )
#dev.off()
# no real improvement without seasonal trend

# Plots - Comparing Out-of-Service Time Series
#png(filename = "Plot - Comparing Out-of-Service Time Series.png")
par(mfrow=c(2,2))
plot(pctCode8 ,type="l",ylab="Out-of-Service Buses",xlab="Days", 
     main = "Daily Percentage of Out-of-Service Buses \n January - April 2015 ", axes = TRUE)
abline(a=meanCode8,b=0, col = 'red',lty="dashed") # comparison line
text(sprintf("Average Daily Percentage \n of Out-of-Service Buses \n %.02f percent", meanCode8),x = 140, y = 20, col="red",cex=0.7) 
abline(a=meanCode8,b=0, col = 'red',lty="dashed") # comparison line
abline(v=40,col = 'red',lty="dashed") # comparison line
text("Sunday, February 9th 2014", x = 140, y = 2, col="red",cex=0.7 )

plot(ts.pctCode8,type="l",col="black",ylab="Out-of-Service Buses",xlab="Week", axes =FALSE, main = "Out-of-Service Buses Times Series")
axis(1,1:80)
axis(2,seq(0,36,2))
text(sprintf("Average Daily Percentage \n of Out-of-Service Buses \n %.02f percent", meanCode8),x = 20, y = 20, col="red",cex=0.7) 
abline(a=meanCode8,b=0, col = 'red',lty="dashed") # comparison line
abline(v=6.56,col = 'red',lty="dashed") # comparison line
text("Sunday, February 9th 2014", x = 20, y = 2, col="red",cex=0.7 )

#png(filename = "Plot - Comparing Out-of-Service Time Series.png")
#par(mfrow=c(2,1))
#par(mar = rep(2, 4))
#plot(ts.pctCode8.detrend,type="l", col="black",ylab="Out-of-Service Buses",xlab="Week", main = "Out-of-Service Times Series \n without Deterministic Trend", axes =FALSE)
#axis(1,1:80)
#axis(2,seq(0,.36,.02))
#abline(a=meanCode8,b=0, col = 'red',lty="dashed") # comparison line

# Plot - Out-of-Service Times Series without Deterministic & Seasonal Trend
plot(ts.pctCode8.deseasonalized,type="l",col="black",ylab="Out-of-Services",xlab="Week", axes =FALSE, main = "Out-of-Service Times Series \n without Deterministic & Seasonal Trend") 
axis(1,1:80)
axis(2,seq(0,36,2))
text(sprintf("Average Daily Percentage \n of Out-of-Service Buses \n %.02f percent", meanCode8),x = 20, y = 20, col="red",cex=0.7) 
abline(a=meanCode8Deaseasonalized,b=0, col = 'red',lty="dashed") # comparison line
abline(v=6.56,col = 'red',lty="dashed") # comparison line
text("Sunday, February 9th 2014", x = 20, y = 2, col="red",cex=0.7 )

#dev.off()

# check for autoregressive (AR) process
#png(filename = "ACF and PACF plots.png")
par(mfrow=c(2,2))
acf(ts.pctCode8) 
acf(ts.pctCode8.deseasonalized) 

pacf(ts.pctCode8)
pacf(ts.pctCode8.deseasonalized)

#dev.off()
par(mfrow=c(1,1))

# Estimate of AR parameters
ar(ts.pctCode8)
ar(ts.pctCode8.deseasonalized)

# Unit root tests for stationarity
PP.test(ts.pctCode8)
adf.test(ts.pctCode8)

PP.test(ts.pctCode8.deseasonalized)
adf.test(ts.pctCode8.deseasonalized)

# Unit root test p-values
PP.count.pvalues <- try(PP.test(ts.pctCode8)$p.value) #Find the PP test p-value
adf.count.pvalues <- try(adf.test(ts.pctCode8)$p.value)  #Find adf test p-value
PP.count.pvalues
adf.count.pvalues

# ADF/PP test null hypothesis is non-stationarity 
# Unit root tests indicate stationarity 

# If non-stationary, use first difference of out-of-service counts and re-run unit root tests
firstDifcode8Buses <- ts(ts.pctCode8.deseasonalized-as.vector(Lag(pctCode8 ,k=1)))
acf(firstDifcode8Buses,na.action=na.pass)
pacf(firstDifcode8Buses,na.action=na.pass)
#adf.test(firstDifcode8Buses[-1])
#PP.test(firstDifcode8Buses[-1])
```

```{r}
# use auto arima to estimate best arima model 
# auto arima suggests AR1I1MA2 or AR2MA2
auto.arima(ts.pctCode8)

# initial model of choice
## Estimate an AR(1), I(1) MA(2) using arima
arima.ar1i1ma2.pctCode8 <- arima(ts.pctCode8, order = c(1,1,2),
                               include.mean = TRUE)

# comparison models
# Estimate an AR(2), I(1) MA(2) using arima
arima.ar2i1ma2.pctCode8 <- arima(ts.pctCode8, order = c(2,1,2),
                                include.mean = TRUE
)

# Estimate an AR(1), I(1) MA(1) using arima
arima.ar1i1ma1.pctCode8 <- arima(ts.pctCode8, order = c(1,1,1),
                               include.mean = TRUE
)

# Estimate an AR(2), I(1) MA(1) using arima
arima.ar2i1ma1.pctCode8 <- arima(ts.pctCode8, order = c(2,1,1),
                               include.mean = TRUE
)

# Estimate an AR(1)MA(2) using arima
arima.ar1i0ma2.pctCode8 <- arima(ts.pctCode8, order = c(1,0,2),
                               include.mean = TRUE)
# Estimate an AR(2)MA(1) using arima
arima.ar2i0ma1.pctCode8 <- arima(ts.pctCode8, order = c(2,0,1),
                               include.mean = TRUE)

# initial model of choice
# Estimate an AR(2)MA(2) using arima
arima.ar2i0ma2.pctCode8 <- arima(ts.pctCode8, order = c(2,0,2),
                               include.mean = TRUE
)

# Extract goodness of fit results from arima models
se.ar1i1ma2.pctCode8 <- arima.ar1i1ma2.pctCode8$sigma2    # standard error of the regression
aic.ar1i1ma2.pctCode8 <- arima.ar1i1ma2.pctCode8$aic      # Akaike Information Criterion
se.ar2i1ma2.pctCode8 <- arima.ar1i1ma2.pctCode8$sigma2    # standard error of the regression
aic.ar2i1ma2.pctCode8 <- arima.ar1i1ma2.pctCode8$aic      # Akaike Information Criterion
se.ar1i1ma1.pctCode8 <- arima.ar1i1ma1.pctCode8$sigma2    # standard error of the regression
aic.ar1i1ma1.pctCode8 <- arima.ar1i1ma1.pctCode8$aic      # Akaike Information Criterion
se.ar2i1ma1.pctCode8 <- arima.ar2i1ma1.pctCode8$sigma2    # standard error of the regression
aic.ar2i1ma1.pctCode8 <- arima.ar2i1ma1.pctCode8$aic      # Akaike Information Criterion
se.arima.ar1i0ma2.pctCode8<- arima.ar1i0ma2.pctCode8$sigma2 # standard error of the regression
aic.arima.ar1i0ma2.pctCode8<- arima.ar1i0ma2.pctCode8$aic # standard error of the regression
se.arima.ar2i0ma1.pctCode8<- arima.ar2i0ma1.pctCode8$sigma2 # standard error of the regression
aic.arima.ar2i0ma1.pctCode8<- arima.ar2i0ma1.pctCode8$aic # standard error of the regression
se.arima.ar2i0ma2.pctCode8<- arima.ar2i0ma2.pctCode8$sigma2 # standard error of the regression
aic.arima.ar2i0ma2.pctCode8<- arima.ar2i0ma2.pctCode8$aic # standard error of the regression

# all standard errors and aic for arima models
all.arima.se.pctCode8<-c(se.ar1i1ma2.pctCode8, se.ar2i1ma2.pctCode8, se.ar1i1ma1.pctCode8,se.ar2i1ma1.pctCode8, se.arima.ar1i0ma2.pctCode8, se.arima.ar2i0ma1.pctCode8, se.arima.ar2i0ma2.pctCode8)
all.arima.aic.pctCode8<-c(aic.ar1i1ma2.pctCode8, aic.ar2i1ma2.pctCode8, aic.ar1i1ma1.pctCode8, aic.ar2i1ma1.pctCode8, aic.arima.ar1i0ma2.pctCode8, aic.arima.ar2i0ma1.pctCode8, aic.arima.ar2i0ma2.pctCode8)


# all standard errors and aic for ARMA models
all.arma.se.pctCode8<-c(se.arima.ar1i0ma2.pctCode8, se.arima.ar2i0ma1.pctCode8, se.arima.ar2i0ma2.pctCode8)
all.arma.aic.pctCode8<-c(aic.arima.ar1i0ma2.pctCode8, aic.arima.ar2i0ma1.pctCode8, aic.arima.ar2i0ma2.pctCode8)
```

```{r, include=FALSE}
# arima cross validation function
arimaCV <- function(x, order, xreg, include.mean, forward=1, minper=50, method = NULL) {
  require(forecast)
  if (!any(class(x)=="ts")) x <- ts(x)
  n <- length(x)
  mae <- matrix(NA, nrow=n-minper, ncol=forward)
  st <- tsp(x)[1]+(minper-2)   
  for(i in 1:(n-minper)) {
    xshort <- window(x, start=st+(i-minper+1), end=st+i)
    xnext <- window(x, start=st+(i+1), end=min(n, st+(i+forward)))
    xregshort <- window(xreg, start=st+(i-minper+1), end=st+i)
    xregnext <- window(xreg, start=st+(i+1), end=min(n, st+(i+forward)))
    fit <- Arima(xshort, order=order, xreg=xregshort, include.mean=include.mean, method
                 =method)
    fcast <- forecast(fit, h=length(xnext), xreg=xregnext)
    mae[i,1:length(xnext)] <- abs(fcast[['mean']]-xnext)
  }
  colMeans(mae, na.rm=TRUE)
}
```

```{r}
# Attempt at rolling window cross-validation 
# comparing out-of service prediction accuracy 15 hypothetical periods (weeks) forward 
# model of choice
cv.ar1i1ma2.pctCode8 <- arimaCV(x=pctCode8 , order=c(1,1,2), forward=10, method = "ML",
                              xreg = firstDifcode8Buses, include.mean=TRUE, minper=15)

# comparison models
# ar2i1ma2
cv.ar2i1ma2.pctCode8 <- arimaCV(pctCode8 , order=c(2,1,2), forward=10, method = "ML",
                              xreg=firstDifcode8Buses, include.mean=TRUE, minper=15)

# ar1i1ma1
cv.ar1i1ma1.pctCode8 <- arimaCV(pctCode8 , order=c(1,1,1), forward=10, method = "ML",
                              xreg=firstDifcode8Buses, include.mean=TRUE, minper=15)

# ar2i1ma1
cv.ar2i1ma1.pctCode8 <- arimaCV(pctCode8 , order=c(2,1,1), forward=10, method = "ML",
                              xreg=firstDifcode8Buses, include.mean=TRUE, minper=15)

# ar1i1ma1
cv.ar1i0ma2.pctCode8 <- arimaCV(pctCode8 , order=c(1,0,2), forward=10, method = "ML",
                              xreg=firstDifcode8Buses, include.mean=TRUE, minper=15)

# ar2i0ma1
cv.ar2i0ma1.pctCode8 <- arimaCV(pctCode8 , order=c(2,0,1), forward=10, method = "ML",
                              xreg=firstDifcode8Buses, include.mean=TRUE, minper=15)

# model of choice
# ar2i0ma2
cv.ar2i0ma2.pctCode8 <- arimaCV(pctCode8 , order=c(2,0,2), forward=10, method = "ML",
                              xreg=firstDifcode8Buses, include.mean=TRUE, minper=15)

# Plot cross-validation results
# ARIMA and ARMA models
all.arima.cv.pctCode8 <- cbind(cv.ar1i1ma2.pctCode8, cv.ar1i1ma1.pctCode8, cv.ar2i1ma2.pctCode8, cv.ar2i1ma1.pctCode8, cv.ar1i0ma2.pctCode8, cv.ar2i0ma1.pctCode8,cv.ar2i0ma2.pctCode8)
avg.arima.cv.pctCode8<-apply(all.cv.pctCode8, 2, mean)
arima.models <- c("AR1I1MA2", "AR2I1MA2","AR1I1MA1", "AR2I1MA1", "AR1MA2", "AR2MA1", "AR2MA2")
arima.model.comparison.pctCode8<-as.data.frame(cbind(arima.models,round(all.arima.aic.pctCode8,3),round(all.arima.se.pctCode8,3), round(avg.arima.cv.pctCode8,3))) 
names(arima.model.comparison.pctCode8)<-c("models","AIC","SE","MAE") 
arima.model.comparison.pctCode8

# ARMA models only
all.arma.cv.pctCode8 <- cbind(cv.ar1i0ma2.pctCode8, cv.ar2i0ma1.pctCode8,cv.ar2i0ma2.pctCode8)
avg.arma.cv.pctCode8<-apply(all.arma.cv.pctCode8, 2, mean)
arma.models <- c("AR1MA2", "AR2MA1", "AR2MA2")
arma.model.comparison.pctCode8<-as.data.frame(cbind(arma.models,round(all.arma.aic.pctCode8,3),round(all.arma.se.pctCode8,3), round(avg.arma.cv.pctCode8,3))) 
names(arma.model.comparison.pctCode8)<-c("models","AIC","SE","MAE") 
arma.model.comparison.pctCode8

col1 <- c("Red", "Yellow", "Blue", "Green", "Orange", "Black", "Purple")
col2 <- c("Blue", "Orange", "Black")

# Plot - ARIMA Cross Validation
#png(filename = "Plot - ARIMA Cross-Validation.png")
matplot(all.cv.pctCode8, type="l", col=col1, lty=1, ylab="Mean Absolute Error", xlab="Periods Forward", main="Cross-validation of ARIMA models", xlim=c(1,13))
text(arima.models, x=rep(12,length(arima.models)), y=all.cv.pctCode8[nrow(all.cv.pctCode8),], col=col)
#dev.off()

# Plot - ARMA Cross Validation
#png(filename = "Plot - ARMA Cross-Validation.png")
matplot(all.arma.cv.pctCode8, type="l", col=col2, lty=1, ylab="Mean Absolute Error", xlab="Periods Forward", main="Cross-validation of ARMA models", xlim=c(1,13))
text(arma.models, x=rep(12,length(arma.models)), y=all.arma.cv.pctCode8[nrow(all.arma.cv.pctCode8),], col=col2)
#dev.off()

# ar1ma2 is best model
# best combination of lowest AIC and MAE
# final model
arma.final.model.pctCode8 <- arima.ar1i0ma2.pctCode8

```

