---
title: 'Linear Model Analysis'
author: 'UW DSSG Paratransit Group'
date: 'updated: Thursday, 2nd July 2015'
output: html_document
---
```{r}
library(timeDate)
library(caret)
library(MASS)
# Make linear regression, regressors vs. total route run time 
setwd("/Users/kivan/repos/DSSG-Paratransit/Access_Analysis_Rproject/reports")
meta_data <- read.csv("../data/ride_meta_data.txt", sep = ",", row.names = NULL)
meta_data <- meta_data[which(meta_data$elapsed_time<86400),]
data <- read.csv("../data/UW_Trip_Data_QC.csv")

# Make matrix of binary contrasts. Every column represents a city, 0 or 1 
# indicates whether a route on a certain day hits a city.
cities <- unique(as.character(data$City))
cities <- cities[cities!=""]
city_mat <- matrix(0, nrow = nrow(meta_data), ncol = length(cities))
colnames(city_mat) <- cities
rownames(city_mat) <- meta_data$run
data$ServiceDate <- as.timeDate(as.character(data$ServiceDate))
data$Run <- as.character(data$Run)
dates <- unique(data$ServiceDate)
runs <- unique(data$Run)
ctr <- 1
for(ii in 1:length(runs)){
  temp <- data[which(data$Run == runs[ii]),]
  temp_days <- unique(temp$ServiceDate)
  for (kk in 1:length(temp_days)){
    this_run <- temp[which(temp$ServiceDate==temp_days[kk]),]
    visited = unique(this_run$City[this_run$City != ""])
    city_mat[ctr,match(visited, cities)] <- 1
    ctr <- ctr +1
  }
}

city_df <- as.data.frame(city_mat)
colnames(city_df) <- cities

# Total run time is a function of maximum clients serviced, average dwell time, and cities hit.
meta_data$max_num_pass[which(meta_data$max_num_pass==-Inf)] = NA

mod1 <- lm(formula = meta_data$elapsed_time ~ meta_data$max_num_pass + meta_data$avg_dwell_time + city_mat)
mod1_summary <- summary(mod1)
#print(mod1_summary)
hist(meta_data$elapsed_time - mod1$fitted.values, main = "Comparison of Elapsed Time with Linear Fit", xlab = " seconds", breaks=seq(-50000,50000,5000))
```

```{r}
#training example
library(mlbench)
data(BostonHousing)
lmFit <- train(medv ~ . + rm:lstat,
data = BostonHousing,
method = "lm")
library(rpart)
rpartFit <- train(medv ~ .,
data = BostonHousing,
method = "rpart",
tuneLength = 9)

plot(rpartFit, compress = TRUE)
```

```{r, eval=FALSE}
seeds <- vector(mode = "list", length = 51)
for(i in 1:50) seeds[[i]] <- sample.int(1000, 22)

## For the last model:
seeds[[51]] <- sample.int(1000, 1)
ctrl <- trainControl(method = "repeatedcv", 
                     repeats = 5,
                     seeds = seeds)
                     
                     
formula <- meta_data$elapsed_time ~ meta_data$max_num_pass + meta_data$avg_dwell_time + city_mat
meta_data$elapsed_time[which(is.nan(meta_data$elapsed_time))] = NA
modelFit <-train(meta_data$elapsed_time ~ meta_data$max_num_pass + meta_data$avg_dwell_time + city_mat,
                 data = meta_data,
                 method ="lm", 
                 trControl=ctrl)

meta_data<-dfr

